Using a Two-Layer Neural Network and Physicochemical Properties to Classify Glass for Forensic Analysis

### 2017-06-30 Paper/paper.tex
Add sensitivity derivation
Add Appendices

\documentclass[12pt,halfline,a4paper]{ouparticle}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[
backend=biber,
style=alphabetic,
sorting=ynt
]{biblatex}
\usepackage[nottoc]{tocbibind}

\addbibresource{paper.bib}

@@

@@

%\abstract{abstract text goes here...}
\date{\today}
\maketitle
\pagebreak
\tableofcontents
\pagebreak

@@

\mathbf{s}^{2} & = \frac{\partial \hat{F}}{\partial \mathbf{n}_{i}^{2}} = - \sum_{s = 1}^{2} \frac{\partial \mathbf{t}_{s} \ln \mathbf{a}_{s}}{\partial \mathbf{n}_{s}} \\
& = - \sum_{s = 1}^{2} \mathbf{t}_{s} \frac{\partial \ln \mathbf{a}_{s}}{\partial \mathbf{n}_{i}} = - \sum_{s = 1}^{2} \mathbf{t}_{s} \frac{1}{\mathbf{a}_{s}} \frac{\partial \mathbf{a}_{s}}{\partial \mathbf{n}_{i}}

@@

where $i$ and $s$ denote the $i$-th and $s$-th neuron, respectively, and 

@@

\frac{\partial a_{s}}{\partial n_{i}} = \dot{\mathbf{F}}^{2}.

@@

\text{if } i = j\text{: } \frac{\partial a_{i}}{\partial n_{i}} = \frac{\partial \textbf{softmax}(n_{i})}{\partial n_{i}}

\text{if } i \neq j\text{: }
\end{split}
\end{equation}

\section{Experimental Setup}
\label{sec5}

\subsection{Data Preprocessing}
\label{sec5.1}

\subsection{Implementation of the Network}
\label{sec5.2}

\subsection{Performance Index}
\label{sec5.3}

@@

\section{Perceptron}
\section{Backprop}
\section{Subroutine X}

>>>>

\documentclass[12pt,halfline,a4paper]{ouparticle}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{listings}
\usepackage{tikz}
\usepackage{fancyhdr}
\usepackage{extramarks}
\usepackage{algpseudocode}
\usetikzlibrary{automata,positioning}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[
backend=biber,
style=alphabetic,
sorting=ynt
]{biblatex}
\usepackage[nottoc]{tocbibind}
\addbibresource{paper.bib}

% defining colors chunks 
\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

% define a global listing style
\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},   
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily,
    breakatwhitespace=false,         
    breaklines=true,                 
    captionpos=b,                    
    keepspaces=true,                 
    numbers=left,                    
    numbersep=3pt,                  
    showspaces=false,                
    showstringspaces=false,
    showtabs=false,                  
    tabsize=1
}

% set global listing style
\lstset{style=mystyle}

% define styles for each language

\lstdefinestyle{R}
    {language = R, style = mystyle}
\lstdefinestyle{Python}
    {language = Python, style = mystyle}
\lstdefinestyle{Matlab}
    {language = Matlab, style = mystyle}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% title page
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%\abstract{abstract text goes here...}
\date{\today}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% table of contents
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\pagebreak
\tableofcontents
\pagebreak

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% paper
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\mathbf{s}^{2} & = \frac{\partial \hat{F}}{\partial \mathbf{n}_{i}^{2}} = - \sum_{i = 1}^{2} \frac{\partial \mathbf{t}_{j} \ln \mathbf{a}_{j}}{\partial \mathbf{n}_{i}} \\
& = - \sum_{j = 1}^{2} \mathbf{t}_{j} \frac{\partial \ln \mathbf{a}_{j}}{\partial \mathbf{n}_{i}} = - \sum_{j = 1}^{2} \mathbf{t}_{j} \frac{1}{\mathbf{a}_{j}} \frac{\partial \mathbf{a}_{j}}{\partial \mathbf{n}_{i}}

@@

where $i$ denotes the $i$-th neuron and 

@@

\frac{\partial a_{i}}{\partial n_{i}} = \dot{\mathbf{F}}^{2}.

@@

\text{if } i = j \text{: } \frac{\partial a_{i}}{\partial n_{i}} & = \frac{\partial \textbf{softmax}(n_{i})}{\partial n_{i}} = \frac{e^{n_{i}}}{\sum_{n = 1}^{2} e^{n_{i}}} \bigg(\frac{\sum_{n = 1}^{2} e^{n_{i}} - e^{n_{i}}}{\sum_{n = 1}^{2} e^{n_{i}}}\bigg) \\ 
& = \frac{e^{n_{i}}}{\sum_{n = 1}^{2} e^{n_{i}}}\bigg(1 - \frac{e^{n_{i}}}{\sum_{n = 1}^{2} e^{n_{i}}}\bigg) = a_{i}(1 - a_{i}) \\ \\
\text{if } i \neq j \text{: } \frac{\partial a_{i}}{\partial n_{j}} & = \frac{\partial \textbf{softmax}(n_{i})}{\partial n_{j}} = - \frac{e^{n_{i}}}{\sum_{n = 1}^{2} e^{n_{i}}} \frac{e^{n_{j}}}{\sum_{n = 1}^{2} e^{n_{j}}} = -a_{i}a_{j}.
Using this result to finish the derivation of $\mathbf{s}^{2}$ from (21), we have: 
s_{i}^{2} &= - \sum_{j = 1}^{2} \mathbf{t}_{j} \frac{1}{\mathbf{a}_{j}} \frac{\partial \mathbf{a}_{j}}{\partial \mathbf{n}_{i}} = - \frac{\mathbf{t}_{i}}{\mathbf{a}_{i}} \frac{\partial \mathbf{a}_{i}}{\partial \mathbf{n}_{i}} - \sum_{j \neq i}^{2} \frac{\mathbf{t}_{j}}{\mathbf{a}_{j}} \frac{\partial \mathbf{a}_{j}}{\partial \mathbf{n}_{i}} \\
& = - \frac{\mathbf{t}_{i}}{\mathbf{a}_{i}} \mathbf{a}_{i}(1 - \mathbf{a}_{i}) - \sum_{j \neq i}^{2} \frac{\mathbf{t}_{j}}{\mathbf{a}_{j}}(-a_{i}a_{j}) \\
& = - \mathbf{t}_{i} + \mathbf{t}_{i}\mathbf{a}_{i} + \sum_{j \neq i}^{2} \mathbf{t}_{j}\mathbf{a}_{i} = - \mathbf{t}_{i} + \sum_{j = 1}^{2} \mathbf{t}_{j}\mathbf{a}_{i} \\
& = - \mathbf{t}_{i} + \mathbf{a}_{i} \sum_{j = 1}^{2} \mathbf{t}_{j} = \mathbf{a}_{i}  - \mathbf{t}_{i}.
Now we have the starting sensitivity: 
\mathbf{s}^{2} = \mathbf{a} - \mathbf{t}
and have everything we need to backpropagate the sensitivities for our network.

\subsubsection{Update Rule}
\label{sec4.2.4}
Lastly, we need to update the weights and biases via an update rule. For the replication in Python, we use the approximate steepest descent rule to update weights and biases: 
\begin{equation}
\mathbf{W}^{m}(k+1) = \mathbf{W}^{m}(k) - \alpha \mathbf{s}^{m}(\mathbf{a}^{m-1})^{T}, 
\end{equation}
\begin{equation}
\mathbf{b}^{m}(k+1) = \mathbf{b}^{m}(k) - \alpha \mathbf{s}^{m}. 
\end{equation}

\section{Experimental Setup}
\label{sec5}
This section will document how the data are used to train and test the network, implementation of the model in Python, and performance metrics used to evaluate the final model.

\subsection{Data Preprocessing}
\label{sec5.1}
The data were almost ready to be used when we downloaded it as there were no missing values. The only thing that needed to be changed from the original dataset was the target values. Originally, these values were coded on a one-to-seven scale where values of one-to-four denoted a type of window glass and values of five-to-seven denoted non-window glass. When coding the targets for pattern recognition problems, it is best practice to have as many dimensions in the targets as there are classes. This ensures only one neuron is active at a time and generally produces better results. Further, we coded the targets as $\begin{bmatrix} 1 \\ 0 \end{bmatrix}$ if the glass sample comes from a window and $\begin{bmatrix} 0 \\ 1 \end{bmatrix}$ it it does not for this reason. We briefly mention this in Section \ref{sec3.2} during our discussion of the targets. The code used to import the data and code the targets can be found in Appendix A.

After collecting the data, we normalize it based on the network architecture we have chosen. Since we use a tangent-sigmoid transfer function in the hidden layer, it is very important that we normalize the inputs between negative one and one since this is the range of the transfer function. When the net input is large (or small), the function becomes saturated. We do not want this to happen at the beginning of the training process because the gradient will be close to zero. Since the first layer's net input is the product of the weight and input plus the bias, large inputs can result in large net inputs if the weights are not small enough scale the inputs down. Thus, it is important to normalize the inputs before we feed them to the network. In order to normalize the inputs to values between negative one and one, we use the following equation: 
\begin{equation}
\mathbf{p}^{n} = 2(\mathbf{p}-\mathbf{p}^{min}) ./ (\mathbf{p}^{max}-\mathbf{p}^{min}) - 1,
\end{equation}
where $\mathbf{p}^{min}$ is the vector containing the minimum values of each element of the input vectors in the dataset, $\mathbf{p}^{max}$ contains the maximum values, ./ represents the element-wise division of the vectors, and $\mathbf{p}^{n}$ is the normalized input vector \cite{hagan2014}.

Next, we divide the data up into three different sets---training, validation, and testing. We chose to use 70\% of the data for training, 15\% of the data for validation, and 15\% of the data for testing. This is fairly standard practice and follows the literature. To ensure that each of these sets is representative of the full dataset, we split the the data up into two subsets---input-target pairs for window samples and input-target pairs for non-windows.

Then, we split each of these subsets up into training, validation, and testing sets using the 70:15:15 ratio. After splitting the subsets up into training, validation, and testing, we combine each split-up subset accordingly---i.e. window-training subset with non-window-training subset. This gives us the highest chance of having our subsets be representative of the full dataset and that the subsets cover the same region of the input space. Once the subsets have been combined into complete training, validation, and testing sets, we reshuffle the input-target pairs so that they are randomly-ordered again. The subroutine used to normalize the inputs and split the data can be found in Appendix B.

\subsection{Implementation of the Network}
\label{sec5.2}
To implement the network, we created a Python class, \verb|NeuronLayer|, to represent layers of a neural network. The \verb|NeuronLayer| class takes the dimension of the inputs, number of neurons in the layer, transfer function, and gradient of the transfer function as arguments. Moreover, there are several functions defined to help train and implement the network in Python. The subroutine used to create the \verb|NeuronLayer| class can be found in Appendix C.

We begin by instantiating layers from the created class to mimic the network architecture we designed. We arbitrarily chose a learning rate, $\alpha$, of 0.01. If the learning rate converges too slowly, then we will increase the learning rate accordingly. Then, we randomly initialize the weights and biases on the interval $[-0.5, 0.5]$ and pass these on to the corresponding network layers with the \verb|setWeightBias| method. Weights and biases are usually initialized on this interval when we have normalized our inputs between negative one and one. Lastly, we initialize empty vectors to hold the error from each training iteration. These steps can be found in Appendix D.

\subsection{Training the Network}
We chose to train the network over 100 epochs where one epoch represents a single batch implementation of approximate steepest descent. The training process begins by propagating each input in the training set forward by calling the \verb|FP| method of the \verb|NeuronLayer| class. Then, the cross-entropy is calculated using the equation found in (8). This error is then used to calculate a sensitivity and then propagated backward through the model. Once this process is done for each input in the dataset, we approximate the cross-entropy loss of the epoch by taking the mean of the cross-entropy of the given epoch. Then, we use this information in addition to the learning rate to update the weights and biases by (26) and (27), respectively.

In attempts to speed up the iterative process used to train the model, we perform the validation simultaneously.  We use the same process to propagate inputs forward and compute the cross-entropy loss, but instead of updating the weights and biases based on the validation set, we use the weights and biases from the training set. This allows us to see if the network is training properly since the validation set is information our model has not been trained on. If the error keeps decreasing in the validation set as the number of iterations increases, this means our network is training properly. To ensure we do not overfit the data with our model, we implement an early stopping technique. We decided to have our network stop training if the validation error increased for five \emph{consecutive} epochs. If this happens to be the case, the script for the model will print to the console that training was stopped early and at what iteration it stopped. The code used to implement this can be found in Appendix E or in the include file \verb|NeuronLayer.py|.

@@

\section{Importing the Data \& Coding the Targets}

The Python code used to import the data and code the targets is listed below. First the data are imported with a standard \verb|read_csv| statement from the \verb|pandas| library. Then, the targets are coded by the process stated in Section 3.2. The full subroutine can be found in the included file \verb|DataPrep.py|.

\begin{lstlisting}[style = Python]
def GlassImport():
    # import dataset
    cols = ['id', 'RI', 'Na', 'Mg', 'Al', 'Si', 'K', 'Ca', 'Ba', 'Fe', 'type']
    glass = pd.read_csv('glass.txt', names=cols)
    # create window target column (1 for true; 0 for false)
    glass.ix[glass.ix[:, 'type'] <= 4, 'window'] = 1
    glass.ix[glass.ix[:, 'type'] > 4, 'window'] = 0
    # create non-window target column (1 for true; 0 for false)
    glass.ix[glass.ix[:, 'type'] > 4, 'non_window'] = 1
    glass.ix[glass.ix[:, 'type'] <= 4, 'non_window'] = 0
    # drop id and class columns
    glass.drop(['id', 'type'], inplace=True, axis=1)
    return glass
\end{lstlisting} 
\pagebreak

\section{Data Normalization \& Division}
The following code documents the process used to normalize the inputs and then divide the data up into training, validation, and testing sets. The full subroutine can be found in the included file \verb|Split.py|.

\begin{lstlisting}[style = Python]
def split(dataset, num_inputs, t, v):
    # normalize dataset into range between -1 and 1
    norm = dataset.ix[:, slice(0,num_inputs)].apply(lambda x: -1 + 2 * (x - x.min()) / (x.max() - x.min()), axis=0)
    dataset = pd.concat([norm, dataset.ix[:,'window':]], axis=1)
    # split dataset by target class
    dataset0 = dataset.ix[dataset.ix[:, 'window'] == 0, :]
    dataset1 = dataset.ix[dataset.ix[:, 'window'] == 1, :]
    # build datasets by target class
    train0, validate0, test0 = np.split(dataset0.sample(frac=1),
                                        [int(t * len(dataset0)), int((1 - v) * len(dataset0))])
    train1, validate1, test1 = np.split(dataset1.sample(frac=1),
                                        [int(t * len(dataset1)), int((1 - v) * len(dataset1))])
    # re-combine by target class
    train = pd.concat([train0, train1])
    validate = pd.concat([validate0, validate1])
    test = pd.concat([test0, test1])
    # shuffle order and reset index numbers
    train = train.sample(frac=1)
    train.reset_index(inplace=True)
    validate = validate.sample(frac=1)
    validate.reset_index(inplace=True)
    test = test.sample(frac=1)
    test.reset_index(inplace=True)
    return train.iloc[:, 1:], validate.iloc[:, 1:], test.iloc[:, 1:]
\end{lstlisting} 
\pagebreak

\section{Python Class for Network Layers}
The following code shows the creation of the \verb|NeuronLayer| class. The full subroutine can be found in the included file \verb|NLObjects.py|.

\begin{lstlisting}[style = Python]
class NeuronLayer:
    # define attributes
    def __init__(self, r, s, f, j):
        self.r = r # Number of inputs
        self.s = s # number of neurons
        self.f = f # Transfer function
        self.j = j # derivatve transfer function

    def setWeightBias(self, w, b):
        self.w = w  # set weights
        self.b = b # set biases
        self.r = w.shape[1]
        self.s = w.shape[0]
        return

    def FP(self, p):
        self.p = p
        self.n = self.w * self.p + self.b
        self.a = self.f(self.n)
        return

    def update(self, learning_rate, sensitivity):
        self.w = self.w - learning_rate * sensitivity * self.p.T
        self.b = self.b - learning_rate * sensitivity
        return
\end{lstlisting} 
\pagebreak

\section{Initializing the Network Architecture in Python}
This code shows the process used to instantiate layers and initialize the parameters of the network.

\begin{lstlisting}[style = Python]
# --------------------- specify network architecture ---------------------
num_neurons1 = 10  # layer 1
num_neurons2 = 2   # layer 2
alpha = 0.01       # learning rate
epoch = 100        # number of iterations
nlayer1 = nl.NeuronLayer(num_neurons1, num_inputs, nl.tansig, nl.j_tansig)      # instantiate layer 1
nlayer2 = nl.NeuronLayer(num_neurons2, num_neurons1, nl.softmax, nl.j_softmax)  # instantiate layer 2
np.random.seed(0)
# randomly initialize weight and bias on the interval [-0.5, 0.5]
W1 = np.matrix(np.random.rand(num_neurons1, num_inputs) - 0.5)
b1 = np.matrix(np.random.rand(num_neurons1, 1) - 0.5)
W2 = np.matrix(np.random.rand(num_neurons2, num_neurons1) - 0.5)
b2 = np.matrix(np.random.rand(num_neurons2, 1) - 0.5)
# initialize cross entropy loss (training & validation)
ce_t = []
ce_v = []
global s1_all, s2_all
# pass on randomly initialized weights and biases to the network
nlayer1.setWeightBias(w=W1, b=b1)
nlayer2.setWeightBias(w=W2, b=b2)
\end{lstlisting} 
\pagebreak

\section{Early Stopping}
The code below shows the early stopping condition used to prevent the model from overfitting.

\begin{lstlisting}[style = Python]
if j == 0:
	val_fail = []
elif ce_v[j] > ce_v[j-1]:
	val_fail.append(1)
	if len(val_fail) == 5:
		print 'Validation error has increased for 5 consecutive epochs. Early stopping at epoch {}'.format(j)
		break
else:
	val_fail = []
\end{lstlisting}
